---
title: 'Week 01: Introduction, nonlinear functions'
author: "Bence Szabó"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: paper
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(error=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(include=TRUE)
knitr::opts_chunk$set(eval=TRUE)
knitr::opts_chunk$set(fig.width=6, fig.height=4)
```

## Introduction

This is a script that creates the outputs for Week 01 of the Econometrics I class, 2023 Fall Semester at Corvinus University of Budapest, Applied Economics.

First we are going to generate plots and data for the Case Study of Life Expectancy and Income. 

## Case Study 01: Life expectancy and Income

```{r}
# Clear memory
rm(list=ls())

# Packages to use
library(tidyverse)
library(modelsummary)
# Estimate piecewise linear splines
library(lspline)
library(fixest)
# For scaling ggplots
require(scales)

```

Load the dataset from the internet.

```{r}
# Call the data from github
my_url <- 'https://osf.io/sh9mu/download'
df     <- read_csv(my_url)
write_csv(df, "week01_worldbank.csv")

df <- df %>% filter(year==2017)

```

```{r}

####
# 
# Good-to-know: Quick check on all HISTOGRAMS
df %>%
  keep(is.numeric) %>% 
  gather() %>% 
  ggplot(aes(value)) +
  facet_wrap(~key, scales = 'free') +
  geom_histogram(bins=20)+
  theme_bw()

datasummary_skim(df)

```


```{r}
######
# Create new variable: Total GDP = GDP per Capita * Populatio
#     note we could have download an other GDP total variable for this,
#     but for comparison purposes, let use the exact same data and 
#     concentrate on difference which are only due to transforming the variables.


df <- df %>% mutate(gdptot = gdppc*population)

```


```{r}

######
# Check basic scatter-plots!
#   Two competing models:
#     A) lifeexp = alpha + beta * gdptot
#     B) lifeexp = alpha + beta * gdppc
#
# Where to use log-transformation? - level-level vs level-log vs log-level vs log-log
# Create the following graphs with loess:


ggplot(df, aes(x = gdptot)) +
  geom_histogram(aes(y=..density..),fill='#3a5e8cFF',color='white') +
  #geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Total GDP (2017 int. const. $, PPP)',y='Percent') +
  scale_y_continuous(labels = scales::percent)+
  theme_bw()

```

```{r}
ggplot(df, aes(x = log(gdptot))) +
  geom_histogram(aes(y=..density..),fill='#3a5e8cFF',color='white') +
  #geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Log of total GDP (2017 int. const. $, PPP)',y='Percent') +
  scale_y_continuous(labels = scales::percent)+
  theme_bw()

```


```{r}

ggplot(df, aes(x = lifeexp)) +
  geom_histogram(aes(y=..density..),fill='#3a5e8cFF',color='white') +
  #geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Life expectancy  (years)',y='Percent') +
  scale_y_continuous(labels = scales::percent)+
  theme_bw()

```

```{r}

ggplot(df, aes(x = log(lifeexp))) +
  geom_histogram(aes(y=..density../100),fill='#3a5e8cFF',color='white') +
  #geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Log of life expectancy  (years)',y='Percent') +
  scale_y_continuous(labels = scales::percent)+
  theme_bw()

```


```{r}
## Model A) lifeexp = alpha + beta * gdptot
# 1) lifeexp - gdptot: level-level model without scaling
ggplot(df, aes(x = gdptot, y = lifeexp)) +
  geom_point(color='#3a5e8cFF') +
  geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Total GDP (2017 int. const. $, PPP)',y = 'Life expectancy  (years)') +
  theme_bw()

```


```{r}
ggplot(df, aes(x =gdptot, y = log(lifeexp))) +
  geom_point(color='#3a5e8cFF') +
  geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Total GDP (2017 int. const. $, PPP)',y = 'Log of Life expectancy  (years)') +
  theme_bw()

ggsave("loglife_gdptot.png", width = 6, height = 4.5)

```

```{r}
ggplot(df, aes(x = log(gdptot), y = log(lifeexp))) +
  geom_point(color='#3a5e8cFF') +
  geom_smooth(method='lm',formula = 'y~x', se = F, color = '#10a53dFF')+
  labs(x = 'Log of Total GDP (2017 int. const. $, PPP)',y = 'Log of Life expectancy  (years)') +
  theme_bw()

```


```{r}
# 2) Change the scale for Total GDP for checking log-transformation
# Tip: you can use `cale_x_continuous(trans = log_trans())` with scales
#   this is good as you can check without creating a new variable
ggplot(df, aes(x = gdptot, y = lifeexp)) +
  geom_point() +
  geom_smooth(method='loess',formula = 'y~x')+
  labs(x = 'Total GDP (2017 int. const. $, PPP, ln scale)',y = 'Life expectancy  (years)') +
  scale_x_continuous(trans = log_trans(),  breaks = c(1,2,5,10,20,50,100,200,500,1000,10000))+
  theme_bw()

```

```{r}
# 3) Change the scale for Total GDP and life-expectancy for checking log-transformation
ggplot(df, aes(x = gdptot, y = lifeexp))  +
  geom_point() +
  geom_smooth(method='loess')+
  labs(x = 'Total GDP (2017 int. const. $, PPP, ln scale)',y = 'Life expectancy  (years, ln scale)') +
  scale_x_continuous(trans = log_trans(),  breaks = c(1,2,5,10,20,50,100,200,500,1000,10000))+
  scale_y_continuous(trans = log_trans())+
  theme_bw()

```


```{r}
###
## Model B) lifeexp = alpha + beta * gdppc:
# 4) lifeexp - gdppc: level-level model without scaling
ggplot(df, aes(x = gdppc, y = lifeexp)) +
  geom_point() +
  geom_smooth(method='loess')+
  labs(x = 'GDP/capita (2017 int. const. $, PPP)',y = 'Life expectancy  (years)') +
  theme_bw()

```

```{r}

# 5) Change the scale for GDP/capita for checking log-transformation
ggplot(df, aes(x = gdppc, y = lifeexp)) +
  geom_point() +
  geom_smooth(method='loess')+
  labs(x = 'GDP/capita (2017 int. const. $, PPP, ln scale)',y = 'Life expectancy  (years)') +
  scale_x_continuous(trans = log_trans(), breaks = seq(0, 120, by = 20))+
  theme_bw()

```

```{r}

# 6) Change the scale for GDP/capita and life-expectancy for checking log-transformation
ggplot(df, aes(x = gdppc, y = lifeexp))  +
  geom_point() +
  geom_smooth(method='loess')+
  labs(x = 'GDP/capita (2017 int. const. $, PPP, ln scale)',y = 'Life expectancy  (years, ln scale)') +
  scale_x_continuous(trans = log_trans(), breaks = seq(0, 120, by = 20)) +
  scale_y_continuous(trans = log_trans()) +
  theme_bw()

```

```{r}
####
# Create new variables: 
#   ln_gdppc  = Log of gdp/capita 
#   ln_gdptot = log GDP total  
# Take Log of gdp/capita and log GDP total
df <- df %>% mutate(gdptot = gdppc * population,
                    gdptot_tn = gdppc * population / 1000, # in trillions
                    ln_gdppc = log(gdppc),
                    ln_gdptot = log(gdptot),
                    ln_gdptot_tn = log(gdptot_tn),
                    ln_lifeexp = log(lifeexp))


######
# Run the following competing models:
#   w ln_gdptot:
#     reg1: lifeexp = alpha + beta * ln_gdptot
#     reg2: lifeexp = alpha + beta_1 * ln_gdptot + beta_2 * ln_gdptot^2
#     reg3: lifeexp = alpha + beta_1 * ln_gdptot + beta_2 * ln_gdptot^2 + beta_3 * ln_gdptot^3
#   w ln_gdppc:
#     reg4: lifeexp = alpha + beta * ln_gdppc
#     reg5: lifeexp = alpha + beta_1 * ln_gdppc + beta_2 * ln_gdppc^2
#     reg6: lifeexp = alpha + beta_1 * ln_gdppc * 1(gdppc < 50) + beta_2 * ln_gdppc * 1(gdppc >= 50)
#   Extra: weighted-ols:
#     reg7: lifeexp = alpha + beta * ln_gdppc, weights: population

###
# Two ways to handle polynomials: 
#
# 1) Add powers of the variable(s) to the dataframe:
df <- df %>% mutate(ln_gdptot_sq = ln_gdptot^2,
                    ln_gdptot_cb = ln_gdptot^3,
                    ln_gdppc_sq  = ln_gdppc^2)
```

```{r }
#
# 2) Use 'poly(x,n)' functions in graphs ONLY, which creates polynomials of x up to order n
#     use this approach for graphs! may use it for models: 
#                   positive - simpler, less new variables, 
#                   negative - uglier names, harder to compare
#     Note: poly() creates rotates your variables automatically to get mean independent variables
#       use raw = TRUE if you dont want to rotate your variables.

# Do the regressions
#
# Using `feols' with classical standard errors
# Reminder: formula: y ~ x1 + x2 + ..., note: intercept is automatically added
reg_b <- feols(lifeexp ~ ln_gdptot_tn, data = df)
reg_b

# Model 0
reg0 <- feols(lifeexp ~ gdptot_tn, data = df, vcov = 'hetero')
reg0

# First model:
reg1 <- feols(lifeexp ~ ln_gdptot_tn, data = df, vcov = 'hetero')
reg1

etable(reg0,reg1,file='gdp_simp_vs_tot.tex')

# Model 0
reg01 <- feols(ln_lifeexp ~ gdptot_tn, data = df, vcov = 'hetero')
reg01
reg01b <- feols(ln_lifeexp ~ gdptot, data = df, vcov = 'hetero')
reg01b
# First model:
reg11 <- feols(ln_lifeexp ~ ln_gdptot_tn, data = df, vcov = 'hetero')
reg11

etable(reg0,
       reg1,
       reg01,
       reg11,
       file='lifeex_gdp_models.tex', replace = TRUE)

```
```{r}

# Visual inspection:
ggplot(data = df, aes(x = ln_gdptot, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(method = lm, color = 'red') +
  theme_bw()

# Second and third model with gdptot
reg2 <- feols(lifeexp ~ ln_gdptot + ln_gdptot_sq, data = df, vcov = 'hetero')
reg2
# Plot
ggplot(data = df, aes(x = ln_gdptot, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(formula = y ~ poly(x,2), method = lm, color = 'red')

reg3 <- feols(lifeexp ~ ln_gdptot + ln_gdptot_sq + ln_gdptot_cb, data = df, vcov = 'hetero')
ggplot(data = df, aes(x = ln_gdptot, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(formula = y ~ poly(x,3), method = lm, color = 'red') +
  theme_bw()

# Compare these models with etable()
etable(reg1, reg2, reg3)
# From these you should consider reg1 and reg3 only!

```

```{r }

##
# Models with gdp per capita:
# reg4: lifeexp = alpha + beta * ln_gdppc
# + plot
reg4 <- feols(lifeexp ~ ln_gdppc, data = df, vcov = 'hetero')
reg4
ggplot(data = df, aes(x = ln_gdppc, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(method = lm, color = 'red')

# reg5: lifeexp = alpha + beta_1 * ln_gdppc + beta_2 * ln_gdppc^2
# + plot
reg5 <- feols(lifeexp ~ ln_gdppc + ln_gdppc_sq, data = df, vcov = 'hetero')
ggplot(data = df, aes(x = ln_gdppc, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(formula = y ~ poly(x,2), method = lm, color = 'red') +
  theme_bw()

##
# Compare results with gdp per capita:
etable(reg4, reg5)
# Conclusion: reg5 is not adding new information

# Compare reg1, reg3 and reg4 to get an idea log transformation is a good idea:
etable(reg1, reg3, reg4)
# R2 measure is much better for reg4...

```

```{r }

##
# Regression with piecewise linear spline:
# 1st define the cutoff for gdp per capita
cutoff <- 50
# 2nd take care of log transformation -> cutoff needs to be transformed as well
# reg6: lifeexp = alpha + beta_1 * ln_gdppc * 1(gdppc < 50) + beta_2 * ln_gdppc * 1(gdppc >= 50)
# + plot
cutoff_ln <- log(cutoff)
# Use simple regression with the lspline function
?lspline
reg6 <- feols(lifeexp ~ lspline(ln_gdppc, cutoff_ln), data = df, vcov = 'hetero')
reg6
ggplot(data = df, aes(x = ln_gdppc, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(formula = y ~ lspline(x,cutoff_ln), method = lm, color = 'red') +
  theme_bw()

ggplot(data = df, aes(x = gdppc, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(formula = y ~ lspline(x, c(cutoff/5, cutoff)), method = lm, color = 'red') +
  theme_bw()

```


```{r }
##
# Extra
# Weighted-OLS: use reg4 setup and weight with population
# Can be done with the `weights = df$population` input!
reg7 <- feols(lifeexp ~ ln_gdppc, data = df, weights = df$population, vcov = 'hetero')
reg7

# Created a pretty graph for visualize this method:
ggplot(data = df, aes(x = ln_gdppc, y = lifeexp)) +
  geom_point(data = df, aes(size=population),  color = 'blue', shape = 16, alpha = 0.6,  show.legend=F) +
  geom_smooth(aes(weight = population), method = 'lm', color='red')+
  scale_size(range = c(1, 15)) +
  coord_cartesian(ylim = c(50, 85)) +
  labs(x = 'ln(GDP per capita, thousand US dollars) ',y = 'Life expectancy  (years)')+
  annotate('text', x = c(4.5, 2.7, 2), y = c(78, 80, 67), label = c('USA', 'China', 'India'), size=5)+
  theme_bw()


#####
# Compare reg4, reg6 and reg7 models with etable:

etable(reg4, reg6, reg7, headers = c('Simple','L.Spline','Weighted'))

#####
# Based on model comparison your chosen model should be reg4 - lifeexp ~ ln_gdppc
#   Substantive: - level-log interpretation works properly for countries
#                - magnitude of coefficients are meaningful
#   Statistical: - simple model, easy to interpret
#                - Comparatively high R2 and captures variation well


######
# Residual analysis.


# Get the predicted y values from the model
df$reg4_y_pred <- reg4$fitted.values
# Calculate the errors of the model
df$reg4_res <- df$lifeexp - df$reg4_y_pred 
# Or alternatively just use reg4$residuals

# Find countries with largest negative errors
worst5 <- df %>% top_n(-5, reg4_res) %>% 
  select(country, lifeexp, reg4_y_pred, reg4_res, ln_gdppc)
worst5

# Find countries with largest positive errors
best5 <- df %>% top_n(5, reg4_res) %>% 
  select(country, lifeexp, reg4_y_pred, reg4_res, ln_gdppc)
best5

# Show again the scatter plot with bests and worst
ggplot(data = df, aes(x = ln_gdppc, y = lifeexp)) + 
  geom_point(color='blue') +
  geom_smooth(method = lm, color = 'red') +
  annotate('text', x = worst5$ln_gdppc, y = worst5$lifeexp - 1, label = worst5$country ,
           color = 'purple') +
  annotate('text', x = best5$ln_gdppc, y = best5$lifeexp + 1, label = best5$country ,
           color = 'green') +
  theme_bw()

```





## Coding session

Let's learn the basics.

















